---
title: "EDA1"
output: html_document
date: "2024-05-16"
---

The script processes a dataset mod_data.csv/mod_r_data.csv, performs data cleaning, transforms categorical and numerical columns, and generates various summary statistics and outputs.

### Prerequisites
R installed on your computer.
Required R packages: dplyr, tidyr, stringr, kableExtra, ggplot2, scales, forcats.

```{r}

library(dplyr)
library(tidyr)
library(stringr)

### Read in dataset
dataset <- read.csv("mod_r_data.csv", header = TRUE, check.names = FALSE)

###Start concat_non_na function: Dataset concatentation function
concat_non_na <- function(x, sep = "-") {
  paste(na.omit(x), collapse = sep)
}
###End concat_non_na function

###Concatenate dataset and add concatenated traces as an end column to the dataset
dataset$Traces <- apply(dataset[,-1], 1, concat_non_na, sep = "-")

###Create new dataset (final_data) comprising Trace ID and concatenated traces
final_data <- dataset %>% select(UniqueID, Traces) 
colnames(final_data)[1] <- "Sequence_ID"

###Calculate each trace length (ex-NA) and add trace lengths as a third column to ###final_data  
final_data$Trace_Length <- sapply(strsplit(final_data$Traces, "-"), function(x) sum(x != "*" & nchar(x) > 0))

###Remove traces with no symbols from the final_data dataset
final_data <- final_data[final_data$Trace_Length != 0, ]

###Count the number of each token types in the dataset 
token_counts <- final_data %>%
  separate_rows(Traces, sep = "-") %>%
  group_by(Traces) %>%
  summarise(Count = n()) %>%
  ungroup() %>%
  arrange(desc(Count)) %>%
  rename(Tokens = Traces)

###Start condense_trace function: This function condenses contiguous sequences of ###one symbol into that symbol with a separator "-" between symbols. Contiguous ###blanks are represented by a "*".
condense_trace <- function(trace) {
  
  trace <- gsub("--", "-*-", trace) # NA -> * placeholder Token
  parts <- unlist(strsplit(trace, "-"))
  condensed <- c()  
  previous <- NULL
  
  for (part in parts) {
    if (part != "" && !identical(part, previous)) {
      condensed <- c(condensed, part)
      previous <- part
    }
  }
  
  condensed_trace <- paste(condensed, collapse = "-")
  
  if (startsWith(condensed_trace, "*") && endsWith(condensed_trace, "*")) {
    if (!startsWith(condensed_trace, "*-")) {
      condensed_trace <- paste0("-*", condensed_trace)
    }
    if (!endsWith(condensed_trace, "-*")) {
      condensed_trace <- paste0(condensed_trace, "-*")
    }
  }
  
  return(condensed_trace)
}
### End condense_trace function

### Adding condensed trace column to final_data
final_data$CT <- sapply(final_data$Traces, condense_trace)

### Calculating the count of symbols in the dataset when treating contiguous 
### sequences of the same symbol as one of that symbol   
Tokens_CT <- unlist(str_split(final_data$CT, "-"))
Token_frequencies_CT <- table(Tokens_CT)
Token_frequencies_CT <- as.data.frame(Token_frequencies_CT)

###Summing the token count for each trace where each sequence of the same symbols ### is treated as one of that symbol and appending as a new column to final_data
final_data$Trace_Length <- sapply(strsplit(final_data$Traces, "-"), function(x) sum(x != "*" & nchar(x) > 0))

###Summing the token count for each trace where each sequence of the same symbols ### is treated as one of that symbol and appending as a new column to final_data
final_data$CT_Length <- sapply(strsplit(final_data$CT, "-"), function(x) sum(x != "*" & nchar(x) > 0))

### Parse sequences and capture all transition pairs (between symbols) in the full ### trace (not the condensed trace) 
transitions_list <- lapply(final_data$Traces, function(sequence) {
  codes <- strsplit(sequence, "-")[[1]]
  # Create pairs of transitions
  pairs <- paste(head(codes, -1), tail(codes, -1), sep = "->")
  return(pairs)
})

### Turn transitions list into a vector
transitions <- unlist(transitions_list)

### Count of all unique transtions types and set out in a dataframe with the symbol ### transition type (from symbol source to symbol target) and count (frequency)
transition_counts <- table(transitions)
transition_df <- as.data.frame(transition_counts, stringsAsFactors = FALSE)
names(transition_df) <- c("Transition", "Frequency")

### Start fix_transition function: This function tidy's up the transition list by ### replacing blanks with "*"
fix_transition <- function(trans) {
  # If '->' is at the start of the string, prepend '*'
  trans <- ifelse(grepl("^->", trans), paste0("*", trans), trans)
  # If '->' is at the end of the string, append '*'
  trans <- ifelse(grepl("->$", trans), paste0(trans, "*"), trans)
  return(trans)
}
### End fix_transition function

### Tidying up transitions usingfix_transition function
transition_df$Transition <- sapply(transition_df$Transition, fix_transition)


### Parse sequences and identify transitions: create a list of each transition for ### each trace where traces are in a list (list of lists)
transitions_list_CT <- lapply(final_data$CT, function(sequence) {
  codes <- strsplit(sequence, "-")[[1]]
  # Create pairs of transitions
  pairs <- paste(head(codes, -1), tail(codes, -1), sep = "->")
  return(pairs)
})

### Vectorise transitions for each trace list of transitions
transitions_CT <- unlist(transitions_list_CT)

```

The script shows the number of rows and columns in the dataset.
The script also displays the number of unique tokens in the dataset.

```{r}

###Please note that an additional column had been added (concatenated traces)
cat("The Dataset has", nrow(dataset), "rows and", ncol(dataset) - 1, "columns. The first column contains a unique trace ID.\n")
cat("\nThe Dataset has", nrow(token_counts), "unique tokens.\n")

```

The following script generates a styled HTML table of token counts.

```{r}

library(kableExtra)

# Generate the table
token_table <- knitr::kable(token_counts, "html") %>%
  kable_styling(full_width = FALSE)

# Add column and row groups
token_table <- kableExtra::column_spec(token_table, 1, width = "30%") %>%
               kableExtra::collapse_rows(columns = 1, valign = "top", latex_hline = "full")

# Output the table
token_table

```

The following code block visualizes token frequency data.

```{r}

library(ggplot2)
library(scales)
library(forcats)

max_count <- max(token_counts$Count)

ggplot(token_counts, aes(x=Tokens, y=Count)) +
  geom_bar(stat="identity", fill="steelblue") +
  theme_minimal() +
  labs(title="Token Frequency", x="Tokens", y="Count") +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 1)
  ) +
  scale_y_continuous(labels = comma, limits=c(0, max_count * 1.2))

token_counts %>% mutate(name = fct_reorder(Tokens, Count)) %>% 
  ggplot(aes(x=name, y=Count)) +
  geom_bar(stat="identity", fill="steelblue") +
  theme_minimal() +
  labs(title="Token Frequency", x="Tokens", y="Count") +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 1)
  ) + coord_flip() +
  scale_y_continuous(labels = comma, limits=c(0, max_count * 1.2))

threshold <- .0001 * max_count

token_counts %>% filter(Count >= threshold)  %>%  mutate(name = fct_reorder(Tokens, Count)) %>%
  ggplot(aes(x=name, y=Count)) +
  geom_bar(stat="identity", fill="steelblue") +
  theme_minimal() +
  labs(title="Token Frequency at 0.01% Threshold of Max Token Count", x="Tokens", y="Count") +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 1)
  ) +  coord_flip() +
  scale_y_continuous(labels = comma, limits=c(0, max_count * 1.2))


```
The following code block visualizes token frequency data for condensed dataset.
The condensed dataset reformats the original data such that traces do not have repeated tokens consecutively (e.g., "A-A-A-A-B-B-B-C-C" becomes "A-B-C")

```{r}

max_count_CT <- max(Token_frequencies_CT$Freq)

ggplot(Token_frequencies_CT, aes(x=Tokens_CT, y=Freq)) +
  geom_bar(stat="identity", fill="steelblue") +
  geom_text(aes(label=scales::comma(Freq)), vjust=-0.1, angle=45, fontface="bold", color="black", size=3.5,nudge_x=0.15) +
  theme_minimal() +
  labs(title="Token Frequency (Condensed Traces)", x="Tokens", y="Count") +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 1)
  ) +
  scale_y_continuous(labels = comma, limits=c(0, max_count_CT * 1.2))

Token_frequencies_CT %>% mutate(name = fct_reorder(Tokens_CT, Freq)) %>% 
  ggplot(aes(x=name, y=Freq)) +
  geom_bar(stat="identity", fill="steelblue") +
  theme_minimal() +
  labs(title="Token Frequency (Condensed Traces)", x="Tokens", y="Count") +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 1)
  ) + coord_flip() +
  scale_y_continuous(labels = comma, limits=c(0, max_count_CT * 1.2))



```

This code block creates two visualizations for the distribution of Trace_Length in the condensed dataset. The condensed dataset reformats the original data such that traces do not have repeated tokens consecutively (e.g., "A-A-A-A-B-B-B-C-C" becomes "A-B-C")

```{r}

ggplot(final_data, aes(x = "", y = CT_Length)) +
  geom_violin(trim = FALSE, fill = "skyblue") +
  geom_boxplot(width = 0.1, fill = "white", color = "black", outlier.shape = NA) +  # Add internal boxplot
  labs(title = "Distribution of Condensed Trace Length",
       y = "Trace Length",
       x = "") +
  theme_minimal() +
  theme(
    axis.title.y = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  coord_flip()

ggplot(final_data, aes(x = CT_Length)) +
  geom_histogram(binwidth = 1, fill = "lightcoral") +
  labs(title = "Distribution of Condensed Trace Length",
       x = "Condensed Trace Length",
       y = "Frequency") +
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )

```

This code block creates two visualizations for the distribution of Trace Length.

```{r}

ggplot(final_data, aes(x = "", y = Trace_Length)) +
  geom_violin(trim = FALSE, fill = "skyblue") +
  geom_boxplot(width = 0.1, fill = "white", color = "black", outlier.shape = NA) +  # Add internal boxplot
  labs(title = "Distribution of Trace Length",
       y = "Trace Length",
       x = "") +
  theme_minimal() +
  theme(
    axis.title.y = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  coord_flip()

ggplot(final_data, aes(x = Trace_Length)) +
  geom_histogram(bins = 20, fill = "skyblue", color = "black") +  # Adjust the number of bins as needed
  labs(title = "Histogram of Trace Length",
       x = "Trace Length",
       y = "Count") +
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )

```

The following code block generates a table of subsequences of traces based on a user-specified length for traces in a condensed dataset.


```{r}

# Sub-sequences 
generate_subsequences <- function(sequence, subseq_length) {
  codes <- unlist(strsplit(sequence, "-"))
  num_codes <- length(codes)
  subsequences <- character(0)
  
  if(num_codes >= subseq_length) {
    for(i in 1:(num_codes - subseq_length + 1)) {
      subsequences <- c(subsequences, paste(codes[i:(i + subseq_length - 1)], collapse = "-"))
    }
  }
  
  return(subsequences)
}

# Request user input for the subsequence length
cat("Please enter the subsequence length (within Condense Trace): ")
subseq_length <- as.integer(readline())

# Validate input
if(is.na(subseq_length) || subseq_length <= 0) {
  stop("Invalid input: subsequence length must be a positive integer.")
}

all_subsequences <- unlist(lapply(final_data$CT, generate_subsequences, subseq_length = subseq_length))
subsequence_counts <- table(all_subsequences)
subsequence_df <- as.data.frame(subsequence_counts, stringsAsFactors = FALSE) %>%
  rename(Subsequence = all_subsequences, Frequency = Freq) %>%
  arrange(desc(Frequency))

# Print the table using kable
kable(subsequence_df, format = "html", caption = "Frequency of Subsequences")
```


The following code block generates a table of subsequences of traces based on a user-specified length for traces in a dataset.

```{r}

# Request user input for the subsequence length
cat("Please enter the subsequence length : ")
ss_length <- as.integer(readline())

# Validate input
if(is.na(ss_length) || ss_length <= 0) {
  stop("Invalid input: subsequence length must be a positive integer.")
}

all_ss <- unlist(lapply(final_data$Traces, generate_subsequences, subseq_length = ss_length))

ss_counts <- table(all_ss)

ss_df <- as.data.frame(ss_counts, stringsAsFactors = FALSE) %>%
  rename(Subsequence = all_ss, Frequency = Freq) %>%
  arrange(desc(Frequency))

# Print the table using kable
kable(ss_df, format = "html", caption = "Frequency of Subsequences")


```

