---
title: "EDA2"
output: html_document
date: "2024-05-20"
---

#In this below code blocks transition token frequency and teh start end token frequency count is analysed and plotted using bubble plot.

Loading the necessary libraries and importing the data.

Combining Sequences: A dataframe dataset_c is created, which combines each of the original dataset's separate                          sequences into a single combined sequence. All of the values in each row of the dataset                           are concatenated to create this combined sequence, with the exception of the first                                column, which is thought to be a unique identifier (UniqueID).


```{r }
library(scales)
library(ggplot2)
library(readr)
library(dplyr)
library(tidyverse)
library(plotly)
library(tidyr)
library(reshape2)


##############Import the dataset################################

dataset <- read.csv("mod_r_data.csv")

#Combine the columns into one column exclusing the UniqueToken column
dataset_c <- data.frame(
  UniqueID = dataset[,1],
  Combined_Sequence = apply(dataset[, -1], 1, function(row) paste(row, collapse = " "))
)

```

#Tranistion token frequncy bubble plot

Token Transitions: For calculating the frequency of transitions between tokens in the concatenated sequences,     the code defines a function called calculate_transition_frequencies. The concatenated sequence is divided        into individual tokens, and the frequency of transitions between tokens is determined. A dataframe called        transition_df is created, containing two columns, Transition and Count, which reflect the frequency of           transitions between tokens and the transition itself, respectively.

Later on the dataframe bubble is created which includes only the transition and the count variables which is then filtered to get the transition without the self transition and the *. 

Finally the plot is build using this filtered_data dataframe which portraits the count as the bubble size and color along with each bubble having the transition token name.

```{r}
# Define a function to calculate transition state frequencies
calculate_transition_frequencies <- function(sequence) {
  tokens <- strsplit(sequence, " ")[[1]]
  transition_frequencies <- table(paste(tokens[-length(tokens)], tokens[-1]))
  transition_df <- data.frame(
    Transition = names(transition_frequencies),
    Count = as.numeric(transition_frequencies),
    stringsAsFactors = FALSE
  )
  return(transition_df)
}

# Apply the function to the Combined_Sequence column
transition_df <- dataset_c %>%
  mutate(Transition_Frequencies = map(Combined_Sequence, calculate_transition_frequencies)) %>%
  unnest(Transition_Frequencies) %>%
  mutate(UniqueID = as.character(UniqueID))

# Pivot the dataframe to wide format
wide_df_transition <- transition_df %>%
  pivot_wider(names_from = Transition, values_from = Count, values_fill = 0)

#All transistion count including self transition 
bubble<-transition_df %>% group_by(Transition) %>% summarise(Count = sum(Count))

# Filter out rows where the transition involves the same token or both tokens are "*"
filtered_data <- bubble %>%
  filter(!grepl("\\b\\* \\*\\b", Transition) & !grepl("\\b(\\w) \\1\\b", Transition))

# Generate random x and y coordinates for bubbles
set.seed(123)  
filtered_data$x_coord <- runif(nrow(filtered_data))
filtered_data$y_coord <- runif(nrow(filtered_data))

# Calculate the range of x and y coordinates
x_range <- range(filtered_data$x_coord)
y_range <- range(filtered_data$y_coord)

# Add a buffer to the range to ensure the bubbles are clipped at the edges
x_buffer <- 0 * diff(x_range)
y_buffer <- 0.12 * diff(y_range)

# Expand the range of x and y coordinates to include the buffer
x_limits <- c(x_range[1] - x_buffer, x_range[2] + x_buffer)
y_limits <- c(y_range[1] - y_buffer, y_range[2] + y_buffer)

# Separate the transition into "From" and "To" tokens
transition_df <- transition_df %>%
  separate(Transition, into = c("From", "To"), sep = " ")

# Calculate the sum of Count for each combination of From and To
transition_matrix <- transition_df %>%
  group_by(From, To) %>%
  summarise(Sum_Count = sum(Count), .groups = "drop")  # Drop grouping information

# Create a full set of combinations of From and To
full_combinations <- expand.grid(From = unique(transition_matrix$From), To = unique(transition_matrix$To))

# Merge with summarized data
transition_matrix <- full_combinations %>%
  left_join(transition_matrix, by = c("From", "To")) %>%
  mutate(Sum_Count = replace_na(Sum_Count, 0))  # Replace missing counts with 0

```


```{r}
# Create a bubble chart with adjusted limits for x and y axes
ggplot(filtered_data, aes(x = x_coord, y = y_coord, size = Count, fill = as.factor(Count))) +
  geom_point(shape = 21, color = "black", alpha = 0.6) +
  scale_size_continuous(range = c(10, 70)) + 
  scale_fill_manual(values = rainbow(length(unique(filtered_data$Count))), guide = "none") +  
  geom_text(aes(label = Transition), size = 3, color = "black", vjust = 0.5) +labs(title = "Transition Count", x = NULL, y = "Transition") + theme_void() + theme(plot.margin = margin(0.1, 1, 0.1, 1, "cm")) + xlim(x_limits) + ylim(y_limits)    

```


```{r}
# Plot the matrix with transparent tiles and light-colored borders based on count
ggplot(data = transition_matrix, aes(x = To, y = From, fill = Sum_Count)) +
  geom_tile(color = "white", size = 1) +  # White borders
  geom_text(aes(label = Sum_Count), color = "black", size = 4) +  
  scale_fill_gradient(low = "lightblue", high = "lightgreen", guide = "none") + 
  labs(x = "To", y = "From", fill = "Count") +
  scale_x_discrete(position = "top") +  
  scale_y_discrete(limits = rev(unique(transition_matrix$To))) +  
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 1))

```


#Start End Token Frequency bubble plot

The process begins by extracting the start and end tokens from the Combined_Sequence column of the dataset.

After extracting the start and end tokens, the next step is to count the occurrences of each unique start and end combination. This provides insights into which combination of start and end tokens are the most common or frequent within the dataset

Once the start and end token along with the occurrence is acquired in this case in the dataframe start_end_df a bubble plot is created where the size and color represent the count and the bubble is labeled with the name of the tokens 


```{r}
#########################Start and End Frequency###############################

# Extract start and end tokens from Combined_Sequence column
start_end <- sapply(dataset_c$Combined_Sequence, function(seq) {
  tokens <- unlist(strsplit(seq, " "))
  
  tokens <- tokens[tokens != "*"]
  
  if (length(tokens) > 1) {
    start <- tokens[1]
    end <- tokens[length(tokens)]
    paste(start, end, sep = " ")
  } else {
    NA
  }
})

start_end_counts <- table(start_end)

start_end_counts <- start_end_counts[!is.na(names(start_end_counts))]

start_end_df <- data.frame(start_end = names(start_end_counts), count = as.vector(start_end_counts))

start_end_df <- start_end_df[order(-start_end_df$count), ] %>% arrange(desc(count))


# Generate random x and y coordinates for bubbles
set.seed(123)  
start_end_df$x_coord <- runif(nrow(start_end_df))
start_end_df$y_coord <- runif(nrow(start_end_df))

```


```{r}
# Create a bubble chart with random coordinates and pastel colors for different sizes
ggplot(start_end_df, aes(x = x_coord, y = y_coord, size = count, fill = as.factor(count))) +
  geom_point(shape = 21, color = "black", alpha = 0.6) +
  scale_size_continuous(range = c(10, 70)) +  
  scale_fill_manual(values = hue_pal()(length(unique(start_end_df$count))), guide = "none") +  
  geom_text(aes(label = start_end), size = 3, color = "black", vjust = 0.5) + 
  labs(title = "Frequency of Start-End Combinations", x = NULL, y = "Frequency") +  
  theme_void() +
  theme(plot.margin = margin(0.1, 1, 0.1, 1, "cm")) 

```
#Single token frequncy boxplot

To determine the frequency of each token in the combined sequences, the code defines the      function calculate_token_frequencies. The combined sequence is divided into separate tokens, and each token's     frequency is counted. A dataframe called token_df is then created, containing two columns called Token and       Count, which stand for the token and its frequency, respectively.

The distribution of the tokens is done using a boxplot where the X axis represents the Tokens and the Y axis represents the Count of that specific token.

```{r}
# Function for calculating the Token Occurences 
calculate_token_frequencies <- function(sequence) {
  # Split the sequence into individual tokens
  tokens <- strsplit(sequence, " ")[[1]]
  # Count the frequency of each token
  token_frequencies <- table(tokens)
  # Create a dataframe with token frequencies
  token_df <- data.frame(
    Token = names(token_frequencies),
    Count = as.numeric(token_frequencies),
    stringsAsFactors = FALSE
  )
  return(token_df)
}

#Calling the above function
token_df <- dataset_c %>%
  mutate(Token_Frequencies = map(Combined_Sequence, calculate_token_frequencies)) %>%
  unnest(Token_Frequencies) %>%  mutate(UniqueID = as.character(UniqueID))


box_plot<-ggplot(token_df, aes(x=Token, y=Count))+ geom_boxplot()+theme_minimal() +
  labs(title = "Distribution of Token Frequencies",
       x = "Token",
       y = "Count") + theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
