---
title: "EDA3"
output: html_document
date: "2024-05-20"
---

The following code block loads necessary libraries for data manipulation, text processing, and graph visualization. It reads a dataset from a CSV file and processes it to create a column of concatenated non-NA traces. It then calculates the length of each trace and removes any traces without valid symbols. The code counts the occurrences of each token type and defines a function to condense contiguous sequences of the same symbol into a single symbol. It adds this condensed trace to the dataset and calculates the frequency of each token in the condensed traces. The code also identifies transitions between symbols in the traces, tidies up the transitions, and creates graph objects to visualize these transitions. Finally, it generates bigrams from the sequences, counts their occurrences, and creates a graph object for the bigrams.


```{r}

library(dplyr)
library(tidyr)
library(stringr)
library(tidytext)
library(tidyverse)
library(igraph)
library(ggraph)

### Read in dataset
dataset <- read.csv("mod_r_data.csv", header = TRUE, check.names = FALSE)

###Start concat_non_na function: Dataset concatentation function
concat_non_na <- function(x, sep = "-") {
  paste(na.omit(x), collapse = sep)
}
###End concat_non_na function

###Concatenate dataset and add concatenated traces as an end column to the dataset
dataset$Traces <- apply(dataset[,-1], 1, concat_non_na, sep = "-")

###Create new dataset (final_data) comprising Trace ID and concatenated traces
final_data <- dataset %>% select(UniqueID, Traces) 
colnames(final_data)[1] <- "Sequence_ID"

###Calculate each trace length (ex-NA) and add trace lengths as a third column to ###final_data  
final_data$Trace_Length <- sapply(strsplit(final_data$Traces, "-"), function(x) sum(x != "*" & nchar(x) > 0))

###Remove traces with no symbols from the final_data dataset
final_data <- final_data[final_data$Trace_Length != 0, ]

###Count the number of each token types in the dataset 
token_counts <- final_data %>%
  separate_rows(Traces, sep = "-") %>%
  group_by(Traces) %>%
  summarise(Count = n()) %>%
  ungroup() %>%
  arrange(desc(Count))

###Start condense_trace function: This function condenses contiguous sequences of ###one symbol into that symbol with a separator "-" between symbols. Contiguous ###blanks are represented by a "*".
condense_trace <- function(trace) {
  
  trace <- gsub("--", "-*-", trace) # NA -> * placeholder Token
  parts <- unlist(strsplit(trace, "-"))
  condensed <- c()  
  previous <- NULL
  
  for (part in parts) {
    if (part != "" && !identical(part, previous)) {
      condensed <- c(condensed, part)
      previous <- part
    }
  }
  
  condensed_trace <- paste(condensed, collapse = "-")
  
  if (startsWith(condensed_trace, "*") && endsWith(condensed_trace, "*")) {
    if (!startsWith(condensed_trace, "*-")) {
      condensed_trace <- paste0("-*", condensed_trace)
    }
    if (!endsWith(condensed_trace, "-*")) {
      condensed_trace <- paste0(condensed_trace, "-*")
    }
  }
  
  return(condensed_trace)
}
### End condense_trace function

### Adding condensed trace column to final_data
final_data$CT <- sapply(final_data$Traces, condense_trace)

### Calculating the count of symbols in the dataset when treating contiguous 
### sequences of the same symbol as one of that symbol   
Tokens_CT <- unlist(str_split(final_data$CT, "-"))
Token_frequencies_CT <- table(Tokens_CT)
Token_frequencies_CT <- as.data.frame(Token_frequencies_CT)

###Summing the token count for each trace where each sequence of the same symbols ### is treated as one of that symbol and appending as a new column to final_data
final_data$CT_Length <- sapply(strsplit(final_data$CT, "-"), function(x) sum(nchar(x) > 0))

### Parse sequences and capture all transition pairs (between symbols) in the full ### trace (not the condensed trace) 
transitions_list <- lapply(final_data$Traces, function(sequence) {
  codes <- strsplit(sequence, "-")[[1]]
  # Create pairs of transitions
  pairs <- paste(head(codes, -1), tail(codes, -1), sep = "->")
  return(pairs)
})

### Turn transitions list into a vector
transitions <- unlist(transitions_list)

### Count of all unique transtions types and set out in a dataframe with the symbol ### transition type (from symbol source to symbol target) and count (frequency)
transition_counts <- table(transitions)
transition_df <- as.data.frame(transition_counts, stringsAsFactors = FALSE)
names(transition_df) <- c("Transition", "Frequency")

### Start fix_transition function: This function tidy's up the transition list by ### replacing blanks with "*"
fix_transition <- function(trans) {
  # If '->' is at the start of the string, prepend '*'
  trans <- ifelse(grepl("^->", trans), paste0("*", trans), trans)
  # If '->' is at the end of the string, append '*'
  trans <- ifelse(grepl("->$", trans), paste0(trans, "*"), trans)
  return(trans)
}
### End fix_transition function

### Tidying up transitions usingfix_transition function
transition_df$Transition <- sapply(transition_df$Transition, fix_transition)


### Parse sequences and identify transitions: create a list of each transition for ### each trace where traces are in a list (list of lists)
transitions_list_CT <- lapply(final_data$CT, function(sequence) {
  codes <- strsplit(sequence, "-")[[1]]
  # Create pairs of transitions
  pairs <- paste(head(codes, -1), tail(codes, -1), sep = "->")
  return(pairs)
})

### Vectorise transitions for each trace list of transitions
transitions_CT <- unlist(transitions_list_CT)

### Count transitions in the vector
transition_counts_CT <- table(transitions_CT)
transition_CT <- as.data.frame(transition_counts_CT, stringsAsFactors = FALSE)
names(transition_CT) <- c("Transition", "Frequency")

### 
transition_df$Source <- sapply(strsplit(as.character(transition_df$Transition), "->"), `[`, 1)
transition_df$Target <- sapply(strsplit(as.character(transition_df$Transition), "->"), `[`, 2)

### Create an igraph graph from the dataframe transition_df
graph <- graph_from_data_frame(d = transition_df[, c("Source", "Target")], directed = TRUE)

### Add edge weights (Frequency)
E(graph)$weight <- transition_df$Frequency

transition_CT$Source <- sapply(strsplit(as.character(transition_CT$Transition), "->"), `[`, 1)
transition_CT$Target <- sapply(strsplit(as.character(transition_CT$Transition), "->"), `[`, 2)

### Create an igraph graph from the dataframe
graph_CT <- graph_from_data_frame(d = transition_CT[, c("Source", "Target")], directed = TRUE)

### Add edge weights (Frequency)
E(graph_CT)$weight <- transition_CT$Frequency


# Function to generate bigrams from a sequence
generate_bigrams <- function(sequence) {
  elements <- strsplit(sequence, "-")[[1]]
  bigrams <- tibble(bigram = paste(head(elements, -1), tail(elements, -1), sep = "-"))
  return(bigrams)
}

# Function to generate bigrams from a sequence
generate_bigrams <- function(sequence) {
  elements <- unlist(strsplit(sequence, "-"))
  bigrams <- tibble(from = head(elements, -1), to = tail(elements, -1))
  return(bigrams)
}

# Apply the function to all sequences and combine the results
bigrams <- final_data %>%
  mutate(CT_bigrams = map(CT, generate_bigrams)) %>%
  unnest(CT_bigrams)

# Count bigram occurrences
bigram_counts <- bigrams %>%
  count(from, to, sort = TRUE)

# Create the graph object
bigram_graph <- graph_from_data_frame(bigram_counts)

```

This code block creates a directed graph to visualize the transition states.

```{r}
layout <- layout_with_fr(graph)

plot(graph, layout = layout,
     edge.arrow.size = 0.3,
     vertex.label.cex = 1,
     vertex.size = sqrt(degree(graph)) * 3,    
     edge.label.cex = 0.3,  
     vertex.color = "green",
     main = "Transition States")
```
This code block creates a directed graph to visualize the transition states (Condensed Traces).

```{r}

layout <- layout_with_fr(graph_CT)

plot(graph_CT, layout = layout,
     edge.arrow.size = 0.3,
     vertex.label.cex = 1,
     vertex.size = sqrt(degree(graph)) * 6,    
     edge.label.cex = 0.3,  
     vertex.color = "green",
     main = "Transition States (Condensed Traces)")

```
This code block visualizes the bigram directed graph.

```{r}

# Arrow specification
a <- grid::arrow(type = "closed", length = unit(.07, "inches"))

# Plot the bigram graph
ggraph(bigram_graph, layout = "fr") +
  geom_edge_link(aes(edge_alpha = n), show.legend = FALSE,
                 arrow = a, end_cap = circle(.07, 'inches')) +
  geom_node_point(color = "lightblue", size = 3) +
  geom_node_text(aes(label = name), vjust = 1, hjust = 1) +
  theme_void() +
  ggtitle("Bigram Directed Graph")

```

