---
title: "Cluster_Model"
output: html_document
date: "2024-05-02"
---

### ClUSTERING MODULE FOR HEATMAP ANLAYSIS

The goal of this module is to do cluster modelling, which includes making cluster plots and finding the ideal k value by using the elbow and silhouette approaches. It also produces summaries of the top five tokens or token transitions. The module provides three different clustering techniques for analysis, which allows for flexibility:

1. K-Means Clustering
2. Hierarchical Clustering
3. Distribution-Based Clustering

The analysis can be conducted on three different types of dataframes:

1. Single token frequency
2. Transitioning token frequency with soft transitions
3. Transition token frequency without soft transitions

Each section of the code block contains instructions preceding the code, for guidence purposes.

***Note: The data set needs to be in the same structure as the mod_r_data dataset in order to work with the below code blocks.***


#### Importing libraries and Dataset

Libraries that are required are loaded in the first code block. Upon loading the necessary libraries, the code uses the readline() function to ask the user to input the directory path containing the dataset. The code uses the setwd() function to set the working directory to the directory path once it has been provided.

Lastly, the code uses the read.csv() method to read the dataset from the designated directory and assigns it to the variable dataset. The dataset with the name "mod_r_data.csv" is presumed to be in CSV format. 

```{r warning=FALSE}
###########################################
#####Importing libraries and Dataset#######
###########################################

library(dplyr)
library(tidyr)
library(tidyverse)
library(ggplot2)
library(stats)
library(cluster)
library(pals)
library(RColorBrewer)
library(data.table)
library(magick) 
library(ComplexHeatmap)###from BioConductor


#Provide the directory path when prompted
dir_path <- readline(prompt = "Enter directory path: ")
#Set the directory path entered 
setwd(dir_path)
# Load dataset
dataset <- read.csv("mod_r_data.csv")#mod_r_dataset.csv

```

#### Preprocessing Datset and creating required Dataframes

This code's goal is to preprocess and convert sequential data into a format that may be used for additional analysis, like clustering. It gathers insightful information regarding the tokens' occurrence and transition patterns inside the sequences, which can be applied to a variety of analytical uses.The tasks that are carried out using the code provided are as below:

-   Combining Sequences: To get started, a new dataframe dataset_c is created, which then combines each of the       original dataset's separate sequences into a single combined sequence. All of the values in each row of the      dataset are concatenated to create this combined sequence, with the exception of the first column, which is      thought to be a unique identifier (UniqueID).

-   Token Occurrences: To determine the frequency of each token in the combined sequences, the code defines the      function calculate_token_frequencies. The combined sequence is divided into separate tokens, and each token's     frequency is counted. A dataframe called token_df is then created, containing two columns called Token and       Count, which stand for the token and its frequency, respectively.

-   Token Transitions: For calculating the frequency of transitions between tokens in the concatenated sequences,     the code defines a function called calculate_transition_frequencies. The concatenated sequence is divided        into individual tokens, and the frequency of transitions between tokens is determined. A dataframe called        transition_df is created, containing two columns, Transition and Count, which reflect the frequency of           transitions between tokens and the transition itself, respectively.

-   Data Transformation: The code uses a map from the purrr package to apply the token frequency and transition      frequency functions to combined sequence in the dataset_c dataframe. The resultant dataframes are then           unnested, pivot_wider is used to convert them into wide format, and the necessary columns are chosen. The        token occurrence and token transition frequencies are found in the final dataframes, data_TO and data_TS,        respectively.

-   The data_TS_wNA dataframe contains a modified version of the data_TS dataframe with the self-transition token     columns removed.

-   Data Cleanup: In order to free up memory, the code uses the rm function to eliminate superfluous intermediary     dataframes (dataset_c, token_df, and transition_df).


```{r}
#################################################################
#####Dataset pre-processing and creating required Dataframes#####
#################################################################

#Combine the columns into one column exclusing the UniqueToken column
dataset_c <- data.frame(
  UniqueID = dataset[,1],
  Combined_Sequence = apply(dataset[, -1], 1, function(row) paste(row, collapse = " "))
)

# Function for calculating the Token Occurences 
calculate_token_frequencies <- function(sequence) {
  # Split the sequence into individual tokens
  tokens <- strsplit(sequence, " ")[[1]]
  # Count the frequency of each token
  token_frequencies <- table(tokens)
  # Create a dataframe with token frequencies
  token_df <- data.frame(
    Token = names(token_frequencies),
    Count = as.numeric(token_frequencies),
    stringsAsFactors = FALSE
  )
  return(token_df)
}

#Calling the above function
token_df <- dataset_c %>%
  mutate(Token_Frequencies = map(Combined_Sequence, calculate_token_frequencies)) %>%
  unnest(Token_Frequencies) %>%  mutate(UniqueID = as.character(UniqueID))

#Pivotting the dataframe from long to wide
data_TO <- token_df %>% 
  pivot_wider(names_from = Token, values_from = Count, values_fill = 0) %>%
  select(-Combined_Sequence)


# Function for counting the frequency of the Token Transitions
calculate_transition_frequencies <- function(sequence) {
  # Split the sequence into individual tokens
  tokens <- strsplit(sequence, " ")[[1]]
  # Initialize transition state frequencies
  transition_frequencies <- table(paste(tokens[-length(tokens)], tokens[-1]))
  # Create a dataframe with transition state frequencies
  transition_df <- data.frame(
    Transition = names(transition_frequencies),
    Count = as.numeric(transition_frequencies),
    stringsAsFactors = FALSE
  )
  return(transition_df)
}

#Calling the tranistion token function
transition_df <- dataset_c %>%
  mutate(Transition_Frequencies = map(Combined_Sequence, calculate_transition_frequencies)) %>%
  unnest(Transition_Frequencies) %>%  mutate(UniqueID = as.character(UniqueID)) # Transition_Frequencies

# Pivot the dataframe to wide format
data_TS <- transition_df %>%
  pivot_wider(names_from = Transition, values_from = Count, values_fill = 0) %>%
  select(-Combined_Sequence)  

#Create the datframe with excluding self transitions
data_TS_wNA <- data_TS %>%   select(-matches("\\*")) %>%   select(-matches("^([A-Z*]) \\1$"))

#Removing the uncessary datasets in the memory
rm(dataset_c, token_df, transition_df)




```

#### Fucnctions Definitions for all analysis

- create_scaled_dataframe: Scales the dataset to prepare it for clustering analysis.

- calculate_and_plot_elbow_method: Calculates the Within-cluster Sum of Squares (WSS) for different values of k    (number of clusters) and plots the results to identify the optimal number of clusters using the Elbow Method.

- calculate_and_plot_silhouette_scores: Computes the Silhouette Score for different values of k and plots the      results to identify the optimal number of clusters based on silhouette analysis.

- perform_kmeans_clustering: Performs k-means clustering on the scaled data with the specified number of clusters   and returns the clustering result.

- create_pca_plot: Creates a PCA plot to visualize the clustering results in reduced dimensions.

- generate_cluster_summary_statistics: Generates summary statistics for each cluster, including token frequencies   and top tokens summary.

- perform_hierarchical_clustering: Performs hierarchical clustering using the Ward's method and returns the        clustering result.

- create_pca_plot_hclust: Creates a PCA plot with hierarchical clustering results to visualize clusters.

- generate_cluster_summary_statistics_hclust: Generates cluster summary statistics for hierarchical clustering     results.

- perform_gmm_clustering: Performs Gaussian Mixture Model (GMM) clustering and returns the clustering result.

- create_pca_plot_gmm: Creates a PCA plot with GMM clustering results to visualize clusters.

- generate_cluster_summary_statistics_gmm: Generates cluster summary statistics for GMM clustering results.

```{r warning=FALSE}
#################################
#####Functions Definitions#####
#################################


# Function to Create Scaled Dataframe
create_scaled_dataframe <- function(data) {
  scaled_data <- data %>%
    select(-1) %>%
    scale()
  return(scaled_data)
}

# Function to Calculate and Plot Elbow Method Results
calculate_and_plot_elbow_method <- function(scaled_data, max_k = 10) {
  set.seed(123)
  wss <- numeric(max_k)
  for (i in 1:max_k) {
    kmeans_model <- kmeans(scaled_data, centers = i)
    wss[i] <- sum(kmeans_model$withinss)
  }
  plot(1:max_k, wss, type = "b", xlab = "Number of Clusters (k)", ylab = "Within-cluster Sum of Squares (wss)",
       main = "Elbow Method")
}

# Function to Calculate and Plot Silhouette Scores
calculate_and_plot_silhouette_scores <- function(scaled_data, max_k = 9) {
  set.seed(123)
  compute_avg_sil <- function(k) {
    model <- kmeans(scaled_data, centers = k, nstart = 25)
    sil <- silhouette(model$cluster, dist(scaled_data))
    mean(sil[, "sil_width"])
  }
  
  avg_sil_width <- sapply(2:max_k, compute_avg_sil)
  plot(2:max_k, avg_sil_width, type='b', xlab="Number of Clusters (k)", ylab="Average Silhouette Width",
       main = "Silhouette Score Method")
  
  optimal_clusters <- which.max(avg_sil_width) + 1
  cat("The optimal number of clusters based on silhouette analysis is", optimal_clusters, 
      "with an average silhouette width of", max(avg_sil_width), "\n")
}

# Function to Perform K-means Clustering
perform_kmeans_clustering <- function(scaled_data, optimal_clusters) {
  set.seed(123)
  kmeans_result <- kmeans(scaled_data, centers = optimal_clusters, nstart = 25)
  return(kmeans_result)
}

# Function to Create PCA Plot
create_pca_plot <- function(scaled_data, kmeans_result) {
  pca_result <- prcomp(scaled_data)
  pca_df <- data.frame(pca_result$x)
  pca_df$cluster <- factor(kmeans_result$cluster)
  ggplot(pca_df, aes(x = PC1, y = PC2, color = cluster)) +
    geom_point() +
    theme_minimal() +
    labs(title = "K-means Clustering with PCA Reduction",
         x = "Principal Component 1",
         y = "Principal Component 2")
}

# Function to Generate Cluster Summary Statistics
generate_cluster_summary_statistics <- function(data, kmeans_result) {
  optimal_k <- length(unique(kmeans_result$cluster))
  for (i in 1:optimal_k) {
    cat(paste("Cluster", i, "summary statistics:\n"))
    cluster_data <- data[kmeans_result$cluster == i, ]
    column_sums <- colSums(cluster_data[, !names(cluster_data) %in% c("UniqueID", "Cluster")])
    cat("Token Frequencies:\n")
    print(column_sums)
    cat("\n")
    top_tokens <- names(sort(column_sums, decreasing = TRUE))[1:5]
    top_tokens_data <- cluster_data[, top_tokens, drop = FALSE]
    cat("\nTop 5 Tokens Summary:\n")
    summary_stats <- summary(top_tokens_data)
    print(summary_stats)
    cat("\n\n")
  }
}

# Function to Generate Cluster Summary Statistics
generate_cluster_summary_statistics_TT <- function(data, kmeans_result) {
  optimal_k <- length(unique(kmeans_result$cluster))
  for (i in 1:optimal_k) {
    cat(paste("Cluster", i, "summary statistics:\n"))
    cluster_data <- data[kmeans_result$cluster == i, ]
    column_sums <- colSums(cluster_data[, !names(cluster_data) %in% c("UniqueID", "Cluster")])
    cat("Transition Frequencies:\n")
    print(column_sums)
    cat("\n")
    top_tokens <- names(sort(column_sums, decreasing = TRUE))[1:5]
    top_tokens_data <- cluster_data[, top_tokens, drop = FALSE]
    cat("\nTop 5 Transitions Summary:\n")
    summary_stats <- summary(top_tokens_data)
    print(summary_stats)
    cat("\n\n")
  }
}


# Function to Perform Hierarchical Clustering
perform_hierarchical_clustering <- function(scaled_data) {
  dist_matrix <- dist(scaled_data)
  hclust_result <- hclust(dist_matrix, method = "ward.D2")
  return(hclust_result)
}

# Function to Create PCA Plot with Hierarchical Clustering Results
create_pca_plot_hclust <- function(scaled_data, hclust_result, k) {
  pca_result <- prcomp(scaled_data)
  pca_df <- data.frame(pca_result$x)
  clusters <- cutree(hclust_result, k = k)
  pca_df$cluster <- factor(clusters)
  ggplot(pca_df, aes(x = PC1, y = PC2, color = cluster)) +
    geom_point() +
    theme_minimal() +
    labs(title = paste("Hierarchical Clustering with PCA Reduction (", k, "clusters)", sep = ""),
         x = "Principal Component 1",
         y = "Principal Component 2")
}

# Function to Generate Cluster Summary Statistics with Hierarchical Clustering Results
generate_cluster_summary_statistics_hclust <- function(data, hclust_result, k) {
  clusters <- cutree(hclust_result, k = k)
  for (i in 1:k) {
    cat(paste("Cluster", i, "summary statistics:\n"))
    cluster_data <- data[clusters == i, ]
    column_sums <- colSums(cluster_data[, !names(cluster_data) %in% c("UniqueID", "Cluster")])
    cat("Token Frequencies:\n")
    print(column_sums)
    cat("\n")
    top_tokens <- names(sort(column_sums, decreasing = TRUE))[1:5]
    top_tokens_data <- cluster_data[, top_tokens, drop = FALSE]
    cat("\nTop 5 Tokens Summary:\n")
    summary_stats <- summary(top_tokens_data)
    print(summary_stats)
    cat("\n\n")
  }
}




# Function to Perform GMM Clustering
perform_gmm_clustering <- function(scaled_data, k) {
  gmm_result <- Mclust(scaled_data, G = k)
  return(gmm_result)
}

# Function to Create PCA Plot with GMM Clustering Results
create_pca_plot_gmm <- function(scaled_data, gmm_result) {
  pca_result <- prcomp(scaled_data)
  pca_df <- data.frame(pca_result$x)
  clusters <- gmm_result$classification
  pca_df$cluster <- factor(clusters)
  ggplot(pca_df, aes(x = PC1, y = PC2, color = cluster)) +
    geom_point() +
    theme_minimal() +
    labs(title = paste("GMM Clustering with PCA Reduction (", length(unique(clusters)), "clusters)", sep = ""),
         x = "Principal Component 1",
         y = "Principal Component 2")
}



# Function to Generate Cluster Summary Statistics for Top Tokens
generate_cluster_summary_statistics_gmm <- function(data, gmm_result) {
  clusters <- gmm_result$classification
  num_clusters <- length(unique(clusters))
  for (i in 1:num_clusters) {
    cat(paste("Cluster", i, "summary statistics:\n"))
    cluster_data <- data[clusters == i, ]
    column_sums <- colSums(cluster_data[, !names(cluster_data) %in% c("UniqueID", "Cluster")])
    cat("Token Frequencies:\n")
    print(column_sums)
    cat("\n")
    top_tokens <- names(sort(column_sums, decreasing = TRUE))[1:5]  # Extract top 5 tokens
    top_tokens_data <- cluster_data[, top_tokens, drop = FALSE]  # Subset data for top tokens
    cat("\nTop 5 Tokens Summary:\n")
    summary_stats <- summary(top_tokens_data)
    print(summary_stats)
    cat("\n\n")
  }
}


```

#### Scaling Data

This code block is dedicated to creating the scaled dataframe calling the fuctions created above.

```{r}

###########################################
##############Scaling Data#################
###########################################

# Create scaled dataframe
  data_TO_scaled <- create_scaled_dataframe(data_TO)

# Create scaled dataframe
  data_TS_scaled <- create_scaled_dataframe(data_TS)

# Create scaled dataframe
  data_TS_wNA_scaled <- create_scaled_dataframe(data_TS_wNA)


```

#### Single Token Occurrence Analysis

-  Users can explore, examine, and visualise clusters in their dataset with this code block's extensive workflow    for clustering analysis through the use of different clustering methods.

-  To assist in determining the ideal number of clusters, the elbow method results are computed and plotted by      the code block.

-  To evaluate the quality of clusters produced silhouette scores are also computed and plotted.

-  The user is prompted to enter the optimal number of clusters.

-  The code carries out the subsequent tasks in accordance with the user's selected clustering option (K-means,     Hierarchical, or Distribution-based):

  *K-means Clustering (Option 1):*
    Performs K-means clustering on the scaled dataset.
    Generates PCA plots to visualize cluster separation.
    Generates cluster summary statistics.
    Visualizes token states by sequence position within each cluster using area plots and heatmaps.
      
  *Hierarchical Clustering (Option 2):*
    Performs hierarchical clustering on the scaled dataset.
    Plots dendrograms to visualize cluster hierarchies.
    Generates cluster summary statistics.
    Visualizes token states by sequence position within each cluster using area plots and heatmaps.
    
  *Distribution-based Clustering (Option 3):*
    Performs distribution-based clustering on the scaled dataset.
    Generates PCA plots to visualize cluster separation.
    Generates cluster summary statistics.
    Visualizes token states by sequence position within each cluster using area plots and heatmaps.
      
***Note: As the library mclust overrides the map function from the purrr package it is loaded separately in necessary code block and removed at the end of the code block***
  
```{r}

###########################################
#####Single Token Occurrence Analysis######
###########################################


library(mclust)


# Plot Elbow Method Results
calculate_and_plot_elbow_method(data_TO_scaled)

# Plot Silhouette Scores
calculate_and_plot_silhouette_scores(data_TO_scaled)

# Determine Optimal Number of Clusters
  optimal_clusters_str <- readline(prompt = "Enter Optimal Cluster Number: ")
  optimal_clusters <- as.numeric(optimal_clusters_str)

option <- as.integer(readline(prompt = "Enter from below which clustering option would you like:\n 1.Kmeans\n 2.Heirarchial\n 3.Distribution "))


switch(option,
       {
         # Perform K-means Clustering
         if (option == 1) {
           kmeans_result <- perform_kmeans_clustering(data_TO_scaled, optimal_clusters)
           plot_TO<-create_pca_plot(data_TO_scaled, kmeans_result)
           print(plot_TO)
           generate_cluster_summary_statistics(data_TO, kmeans_result)
           cluster_labels <- kmeans_result$cluster
           data_TO_Clusters_K<- cbind(data_TO, cluster_labels)
           file_path <- "data_TO_Kmeans_cluster.csv"
           if (file.exists(file_path)) {
              message("File exists. It will be replaced.")
            } else {
              message("File does not exist. A new file will be created.")
            }

           write.csv(data_TO_Clusters_K, file = file_path, row.names = FALSE)
         }
       },
       {
         # Perform Heirarchical clustering
         if (option == 2) {
           hclust_result_TO <- perform_hierarchical_clustering(data_TO_scaled)
           plot(hclust_result_TO, hang = -1, cex = 0.6, main = "Dendrogram")
           k <- optimal_clusters
           plot_TO<-create_pca_plot_hclust(data_TO_scaled, hclust_result_TO, k)
           print(plot_TO)
           generate_cluster_summary_statistics_hclust(data_TO, hclust_result_TO, k)
            # Extract cluster labels from hclust result
            cluster_labels <- cutree(hclust_result_TO, k = k)
            data_TO_Clusters_h <- data_TO
            data_TO_Clusters_h$cluster_labels <- cluster_labels
            file_path <- "data_TO_Hierarchical_cluster.csv"
            if (file.exists(file_path)) {
               message("File exists. It will be replaced.")
             } else {
               message("File does not exist. A new file will be created.")
             }

           write.csv(data_TO_Clusters_h, file = file_path, row.names = FALSE)

         }
       },
       {
         # Perform distribution-based clustering
         if (option == 3) {
           k <- optimal_clusters  # Adjust the number of components as needed
           gmm_result_TO <- perform_gmm_clustering(data_TO_scaled, k)
           plot_TO<-create_pca_plot_gmm(data_TO_scaled, gmm_result_TO)
           print(plot_TO)
           generate_cluster_summary_statistics_gmm(data_TO, gmm_result_TO)
           cluster_labels <- gmm_result_TO$classification
           data_TO_Clusters_d<-cbind(data_TO, cluster_labels)
           file_path <- "data_TO_Distribution_cluster.csv"
           if (file.exists(file_path)) {
               message("File exists. It will be replaced.")
             } else {
               message("File does not exist. A new file will be created.")
             }

           write.csv(data_TO_Clusters_d, file = file_path, row.names = FALSE)

         }
       },
       {
         # Default case
         cat("Invalid option selected. Please select 1, 2, or 3.")
       }
)


if ("cluster" %in% colnames(dataset)) {
  # If it exists, delete the 'cluster_labels' column
  dataset <- dataset[, -which(colnames(dataset) == "cluster")]
}

# Rename columns, first stays original, rest are numbered sequentially
col_num_name <- seq(1, ncol(dataset)-1, by=1)
col_num_name <- c(names(dataset[1]), col_num_name)
colnames(dataset) <- col_num_name

# Convert the dataset to a matrix and then to a vector, extracting unique columns
unique_tokens <- unique(as.vector(as.matrix(dataset[,2:ncol(dataset)])))

# Assign high-contrast colors to each unique token for heatmap coloring
heatmap_color <- structure(kelly(n=length(unique_tokens)), names = unique_tokens)

print(unique_tokens)

switch(option,
       {
         # Perform K-means Clustering
         if (option == 1) {

            dataset$cluster <- data_TO_Clusters_K$cluster_labels
            
            for (i in 1:max(dataset$cluster)){
            
              # Subsetting dataset by cluster
              data_clust <- dataset[dataset$cluster == i, ]  
              
              # Remove last column
              data_clust$cluster <- NULL
              
              # Aggregating by token count at each point in the sequence 
              token_states <- data_clust %>% gather(key, value, -UniqueID) %>% dplyr::select(-UniqueID) %>%                     table()                        
            
              # Turning token_states into a dataframe...
              state_map <- as.data.frame.matrix(token_states) 
            
              # Creating a rowname vector and ordering on this (rownames are characters - so convert) 
              rn <- rownames(state_map)
              state_map <- state_map[order(as.numeric(rn)), ]
              
              # Reorders columns
              state_map <- state_map[,order(colSums(-state_map,na.rm=TRUE))]
            
              allnames <- colnames(state_map)
            
              # Creating x-axis numeric label
              state_map$SeqPosition <- as.numeric(rownames(state_map))
            
              # Turn state-map into a data table
              state_map <- setDT(state_map) 
            
              state_final <- melt(state_map, id.vars = "SeqPosition", measure.vars = allnames)
              
              state_final <- rename(state_final, "Token" = "variable")
              
              TS_title <-paste0("Count of token states by sequence position - Cluster: ", as.character(i)) 
            
              # Print state plot
              print(ggplot(state_final, aes(x=SeqPosition, y=value, fill=Token)) + scale_fill_manual(values                     = heatmap_color) + geom_area() + ylab("Token Count") + xlab("Sequence Position") +                                ggtitle(TS_title) + theme(legend.position="bottom"))
              
              # Turning the dataframe into a matrix to use in complex heatmap 
              heatmap_matrix <- as.matrix(data_clust[, 2:(ncol(dataset)-1)])
            
              # Setting row/col names to NULL, not necessary
              colnames(heatmap_matrix) = NULL
              rownames(heatmap_matrix) = NULL
              
              # This eliminates an option message
              ht_opt$message = FALSE 
            
              HM_title <-paste0("Token Heatmap Cluster: ", as.character(i))
              
              # Printing heatmap
              print(Heatmap(heatmap_matrix, name = "mat", col = heatmap_color, column_title = HM_title))
            }

          }
            
         },
         {
          if (option == 2) {
            dataset$cluster <- data_TO_Clusters_h$cluster_labels

            for (i in 1:max(dataset$cluster)){
            
              # Subsetting dataset by cluster
              data_clust <- dataset[dataset$cluster == i, ]  
              
              # Remove last column
              data_clust$cluster <- NULL
              
              # Aggregating by token count at each point in the sequence 
              token_states <- data_clust %>% gather(key, value, -UniqueID) %>% dplyr::select(-UniqueID) %>%                     table()                        
            
              # Turning token_states into a dataframe...
              state_map <- as.data.frame.matrix(token_states) 
            
              # Creating a rowname vector and ordering on this (rownames are characters - so convert) 
              rn <- rownames(state_map)
              state_map <- state_map[order(as.numeric(rn)), ]
              
              # Reorders Column
              state_map <- state_map[,order(colSums(-state_map,na.rm=TRUE))]
            
              allnames <- colnames(state_map)
            
              # Creating x-axis numeric label
              state_map$SeqPosition <- as.numeric(rownames(state_map))
            
              # Turn state-map into a data table
              state_map <- setDT(state_map) 
            
              state_final <- melt(state_map, id.vars = "SeqPosition", measure.vars = allnames)
              
              state_final <- rename(state_final, "Token" = "variable")
              
              TS_title <-paste0("Count of token states by sequence position - Cluster: ", as.character(i)) 
            
              # Print state plot
              print(ggplot(state_final, aes(x=SeqPosition, y=value, fill=Token)) + scale_fill_manual(values                     = heatmap_color) + geom_area() + ylab("Token Count") + xlab("Sequence Position") +                                ggtitle(TS_title) + theme(legend.position="bottom"))
              # Turning the dataframe into a matrix to use in complexheatmap 
              heatmap_matrix <- as.matrix(data_clust[, 2:(ncol(dataset)-1)]) 
            
              # Setting row/col names to NULL, not necessary
              colnames(heatmap_matrix) = NULL
              rownames(heatmap_matrix) = NULL
              
              # This eliminates an option message
              ht_opt$message = FALSE 
            
              HM_title <-paste0("Token Heatmap Cluster: ", as.character(i))
              
              # Printing heatmap
              print(Heatmap(heatmap_matrix, name = "mat", col = heatmap_color, column_title = HM_title))
            }
          }
        },
           {
             if (option == 3) {
                dataset$cluster <- data_TO_Clusters_d$cluster_labels
                for (i in 1:max(dataset$cluster)){
                  data_clust <- dataset[dataset$cluster == i, ]  
                  data_clust$cluster <- NULL
                  token_states <- data_clust %>% gather(key, value, -UniqueID) %>% dplyr::select(-UniqueID)                   %>% table()    
                  state_map <- as.data.frame.matrix(token_states) 
                  rn <- rownames(state_map)
                  state_map <- state_map[order(as.numeric(rn)), ]
                  state_map <- state_map[,order(colSums(-state_map,na.rm=TRUE))]
                  allnames <- colnames(state_map)
                  state_map$SeqPosition <- as.numeric(rownames(state_map))
                  state_map <- setDT(state_map)
                  state_final <- melt(state_map, id.vars = "SeqPosition", measure.vars = allnames)
                  state_final <- rename(state_final, "Token" = "variable")
                  TS_title <-paste0("Count of token states by sequence position - Cluster: ",                                      as.character(i)) 
                  print(ggplot(state_final, aes(x=SeqPosition, y=value, fill=Token)) +                                             scale_fill_manual(values = heatmap_color) + geom_area() + ylab("Token Count") +                                  xlab("Sequence Position") + ggtitle(TS_title) + theme(legend.position="bottom"))
                  
                  #Turning the dataframe into a matrix to use in complexheatmap 
                  heatmap_matrix <- as.matrix(data_clust[, 2:(ncol(dataset)-1)]) 
                
                  # Setting row/col names to NULL, not necessary
                  colnames(heatmap_matrix) = NULL
                  rownames(heatmap_matrix) = NULL
                  
                  # This eliminates an option message
                  ht_opt$message = FALSE 
                
                  HM_title <-paste0("Token Heatmap Cluster: ", as.character(i))
                  
                  # Printing heatmap
                  print(Heatmap(heatmap_matrix, name = "mat", col = heatmap_color, column_title = HM_title))
                }
               
              }
           }
)
         

detach("package:mclust")

```

#### Clustering Analysis with Transitioning Tokens (Including Self-Transitions)


This code block operates similarly to the previous one but focuses on the dataframe containing transitioning tokens, including self-transitions. The flow of this code block mirrors that of the preceding one, with the key distinction being the dataset used. Despite this difference, the overall process remains consistent with previous iterations (code block 5).

```{r  warning=FALSE}
###################################################################
#####Transitioning Tokens analsis (Including Self-Transitions)#####
###################################################################

library(mclust)

# Plot Elbow Method Results
  calculate_and_plot_elbow_method(data_TS_scaled)

# Plot Silhouette Scores
  calculate_and_plot_silhouette_scores(data_TS_scaled)


# Determine Optimal Number of Clusters
  optimal_clusters_str <- readline(prompt = "Enter Optimal Cluster Number: ")
  optimal_clusters <- as.numeric(optimal_clusters_str)

option <- as.integer(readline(prompt = "Enter from below which clustering option would you like:\n 1.Kmeans\n 2.Heirarchial\n 3.Distribution "))

switch(option,
       {
         # Perform K-means Clustering
         if (option == 1) {

            # Perform K-means Clustering
              kmeans_result <- perform_kmeans_clustering(data_TS_scaled, optimal_clusters)
              plot_TS<-create_pca_plot(data_TS_scaled, kmeans_result)
              print(plot_TS)
              generate_cluster_summary_statistics_TT(data_TS, kmeans_result)
              cluster_labels <- kmeans_result$cluster
              data_TS_Clusters_k<- cbind(data_TS, cluster_labels)
              file_path <- "data_TS_Kmeans_cluster.csv"
             if (file.exists(file_path)) {
               message("File exists. It will be replaced.")
             } else {
               message("File does not exist. A new file will be created.")
             }

           write.csv(data_TS_Clusters_k, file = file_path, row.names = FALSE)

         }
       },
       {
         # Perform Heirarchical
         if (option == 2) {
  
            hclust_result_TS <- perform_hierarchical_clustering(data_TS_scaled)
            plot(hclust_result_TS, hang = -1, cex = 0.6, main = "Dendrogram")
            k<-optimal_clusters
            plot_TS<-create_pca_plot_hclust(data_TS_scaled, hclust_result_TS, k)
            print(plot_TS)
            generate_cluster_summary_statistics_hclust(data_TS, hclust_result_TS, k)
            cluster_labels <- cutree(hclust_result_TS, k = k)
            data_TS_Clusters_h <- data_TS
            data_TS_Clusters_h$cluster_labels <- cluster_labels
            file_path <- "data_TS_Hierarchical_cluster.csv"
            if (file.exists(file_path)) {
               message("File exists. It will be replaced.")
            } else {
               message("File does not exist. A new file will be created.")
            }

           write.csv(data_TS_Clusters_h, file = file_path, row.names = FALSE)

         }
       },
       {
         if (option == 3) {

          k <- optimal_clusters
          gmm_result_TS <- perform_gmm_clustering(data_TS_scaled, k)
          plot_TS<-create_pca_plot_gmm(data_TS_scaled, gmm_result_TS)
          print(plot_TS)
          generate_cluster_summary_statistics_gmm(data_TS, gmm_result_TS)
          cluster_labels <- gmm_result_TS$classification
          data_TS_Clusters_d<-cbind(data_TS, cluster_labels)
          file_path <- "data_TS_Distribution_cluster.csv"
          if (file.exists(file_path)) {
             message("File exists. It will be replaced.")
          } else {
             message("File does not exist. A new file will be created.")
          }

           write.csv(data_TS_Clusters_d, file = file_path, row.names = FALSE)

         }
       },
       {
         # Default case
         cat("Invalid option selected. Please select 1, 2, or 3.")
       }

)



if ("cluster" %in% colnames(dataset)) {
  # If it exists, delete the 'cluster_labels' column
  dataset <- dataset[, -which(colnames(dataset) == "cluster")]
}

# Rename columns, first stays original, rest are numbered sequentially
col_num_name <- seq(1, ncol(dataset)-1, by=1)
col_num_name <- c(names(dataset[1]), col_num_name)
colnames(dataset) <- col_num_name

# Convert the dataset to a matrix and then to a vector, extracting unique columns
unique_tokens <- unique(as.vector(as.matrix(dataset[,2:ncol(dataset)])))

# Assign high-contrast colors to each unique token for heatmap coloring
heatmap_color <- structure(kelly(n=length(unique_tokens)), names = unique_tokens)

print(unique_tokens)

switch(option,
       {
         # Perform K-means Clustering
         if (option == 1) {

            dataset$cluster <- data_TS_Clusters_k$cluster_labels
            
            for (i in 1:max(dataset$cluster)){
            
              # Subsetting dataset by cluster
              data_clust <- dataset[dataset$cluster == i, ]  
              
              # Remove last column
              data_clust$cluster <- NULL
              
              # Aggregating by token count at each point in the sequence 
              token_states <- data_clust %>% gather(key, value, -UniqueID) %>% dplyr::select(-UniqueID) %>%               table()                        
            
              # Turning token_states into a dataframe...
              state_map <- as.data.frame.matrix(token_states) 
            
              # Creating a rowname vector and ordering on this (rownames are characters - so convert) 
              rn <- rownames(state_map)
              state_map <- state_map[order(as.numeric(rn)), ]
              
              # Reorders columns
              state_map <- state_map[,order(colSums(-state_map,na.rm=TRUE))]
            
              allnames <- colnames(state_map)
            
              # Creating x-axis numeric label
              state_map$SeqPosition <- as.numeric(rownames(state_map))
            
              # Turn state-map into a data table
              state_map <- setDT(state_map) 
            
              state_final <- melt(state_map, id.vars = "SeqPosition", measure.vars = allnames)
              
              state_final <- rename(state_final, "Token" = "variable")
              
              TS_title <-paste0("Count of token states by sequence position - Cluster: ", as.character(i)) 
            
              # Print state plot
              print(ggplot(state_final, aes(x=SeqPosition, y=value, fill=Token)) + scale_fill_manual(values                    = heatmap_color) + geom_area() + ylab("Token Count") + xlab("Sequence Position") +                               ggtitle(TS_title) + theme(legend.position="bottom"))
              # Turning the dataframe into a matrix to use in complexheatmap 
              heatmap_matrix <- as.matrix(data_clust[, 2:(ncol(dataset)-1)]) 
            
              # Setting row/col names to NULL, not necessary
              colnames(heatmap_matrix) = NULL
              rownames(heatmap_matrix) = NULL
              
              # This eliminates an option message
              ht_opt$message = FALSE 
            
              HM_title <-paste0("Token Heatmap Cluster: ", as.character(i))
              
              # Printing heatmap
              print(Heatmap(heatmap_matrix, name = "mat", col = heatmap_color, column_title = HM_title))
            }

          }
            
         },
         {
          if (option == 2) {
            dataset$cluster <- data_TS_Clusters_h$cluster_labels

            for (i in 1:max(dataset$cluster)){
            
              # Subsetting dataset by cluster
              data_clust <- dataset[dataset$cluster == i, ]  
              
              # Remove last column
              data_clust$cluster <- NULL
              
              # Aggregating by token count at each point in the sequence 
              token_states <- data_clust %>% gather(key, value, -UniqueID) %>% dplyr::select(-UniqueID) %>%               table()                        
            
              # Turning token_states into a dataframe...
              state_map <- as.data.frame.matrix(token_states) 
            
              # Creating a rowname vector and ordering on this (rownames are characters - so convert) 
              rn <- rownames(state_map)
              state_map <- state_map[order(as.numeric(rn)), ]
              
              # Reorders columns
              state_map <- state_map[,order(colSums(-state_map,na.rm=TRUE))]
            
              allnames <- colnames(state_map)
            
              # Creating x-axis numeric label
              state_map$SeqPosition <- as.numeric(rownames(state_map))
            
              # Turn state-map into a data table
              state_map <- setDT(state_map) 
            
              state_final <- melt(state_map, id.vars = "SeqPosition", measure.vars = allnames)
              
              state_final <- rename(state_final, "Token" = "variable")
              
              TS_title <-paste0("Count of token states by sequence position - Cluster: ", as.character(i)) 
            
              # Print state plot
              print(ggplot(state_final, aes(x=SeqPosition, y=value, fill=Token)) + scale_fill_manual(values                    = heatmap_color) + geom_area() + ylab("Token Count") + xlab("Sequence Position") +                               ggtitle(TS_title) + theme(legend.position="bottom"))
              
              # Turning the dataframe into a matrix to use in complexheatmap 
              heatmap_matrix <- as.matrix(data_clust[, 2:(ncol(dataset)-1)]) 
            
              # Setting row/col names to NULL, not necessary
              colnames(heatmap_matrix) = NULL
              rownames(heatmap_matrix) = NULL
              
              # This eliminates an option message
              ht_opt$message = FALSE 
            
              HM_title <-paste0("Token Heatmap Cluster: ", as.character(i))
              
              # Printing heatmap
              print(Heatmap(heatmap_matrix, name = "mat", col = heatmap_color, column_title = HM_title))
            }
          }
        },
           {
             if (option == 3) {
                dataset$cluster <- data_TS_Clusters_d$cluster_labels
                for (i in 1:max(dataset$cluster)){
                  data_clust <- dataset[dataset$cluster == i, ]  
                  data_clust$cluster <- NULL
                  token_states <- data_clust %>% gather(key, value, -UniqueID) %>% dplyr::select(-UniqueID)                         %>% table()    
                  state_map <- as.data.frame.matrix(token_states) 
                  rn <- rownames(state_map)
                  state_map <- state_map[order(as.numeric(rn)), ]
                  state_map <- state_map[,order(colSums(-state_map,na.rm=TRUE))]
                  allnames <- colnames(state_map)
                  state_map$SeqPosition <- as.numeric(rownames(state_map))
                  state_map <- setDT(state_map)
                  state_final <- melt(state_map, id.vars = "SeqPosition", measure.vars = allnames)
                  state_final <- rename(state_final, "Token" = "variable")
                  TS_title <-paste0("Count of token states by sequence position - Cluster: ",                                      as.character(i)) 
                  print(ggplot(state_final, aes(x=SeqPosition, y=value, fill=Token)) +                                             scale_fill_manual(values = heatmap_color) + geom_area() + ylab("Token Count") +                                  xlab("Sequence Position") + ggtitle(TS_title) + theme(legend.position="bottom"))
               
                  # Turning the dataframe into a matrix to use in complex heatmap 
                  heatmap_matrix <- as.matrix(data_clust[, 2:(ncol(dataset)-1)]) 
                
                  # Setting row/col names to NULL, not necessary
                  colnames(heatmap_matrix) = NULL
                  rownames(heatmap_matrix) = NULL
                  
                  # This eliminates an option message
                  ht_opt$message = FALSE 
                
                  HM_title <-paste0("Token Heatmap Cluster: ", as.character(i))
                  
                  # Printing heatmap
                  print(Heatmap(heatmap_matrix, name = "mat", col = heatmap_color, column_title = HM_title))
                }
               
              }
           }
)
  





detach("package:mclust")

```
#### Clustering Analysis with Transitioning Tokens (Excluding Self-Transitions)

Similar to the preceding code blocks, this section will analyze the dataframe containing transitioning tokens, excluding self-transitions. The process entails performing the same operations as described in the code block 5, but specifically focused on this subset of data.

```{r warning=FALSE}

###################################################################
#####Transitioning Tokens analsis (Excluding Self-Transitions)#####
###################################################################

library(mclust)

# Plot Elbow Method Results
  calculate_and_plot_elbow_method(data_TS_wNA_scaled)
  
# Plot Silhouette Scores
  calculate_and_plot_silhouette_scores(data_TS_wNA_scaled)
  
# Determine Optimal Number of Clusters
  optimal_clusters_str <- readline(prompt = "Enter Optimal Cluster Number: ")
  optimal_clusters <- as.numeric(optimal_clusters_str)
  
option <- as.integer(readline(prompt = "Enter from below which clustering option would you like:\n 1.Kmeans\n 2.Heirarchial\n 3.Distribution "))

switch(option,
       {
         # Perform K-means Clustering
         if (option == 1) {
  
          # Perform K-means Clustering
            kmeans_result <- perform_kmeans_clustering(data_TS_wNA_scaled, optimal_clusters)
            plot_TS_wNA<-create_pca_plot(data_TS_wNA_scaled, kmeans_result)
            print(plot_TS_wNA)
            generate_cluster_summary_statistics_TT(data_TS_wNA, kmeans_result)
            cluster_labels <- kmeans_result$cluster
            data_TS_wNA_Clusters_k<- cbind(data_TS_wNA, cluster_labels)
            file_path <- "data_TS_wNA_Kmeans_cluster.csv"
            if (file.exists(file_path)) {
               message("File exists. It will be replaced.")
            } else {
               message("File does not exist. A new file will be created.")
            }

           write.csv(data_TS_wNA_Clusters_k, file = file_path, row.names = FALSE)

         }
       },
       {
         # Perform Heirarchical
         if (option == 2) {
            hclust_result_wNA <- perform_hierarchical_clustering(data_TS_wNA_scaled)
            plot(hclust_result_wNA, hang = -1, cex = 0.6, main = "Dendrogram")
            k<-optimal_clusters
            plot_TS_wNA<-create_pca_plot_hclust(data_TS_wNA_scaled, hclust_result_wNA, k)
            print(plot_TS_wNA)
            generate_cluster_summary_statistics_hclust(data_TS_wNA, hclust_result_wNA, k)
            cluster_labels <- cutree(hclust_result_wNA, k = k)
            data_TS_wNA_Clusters_h <- data_TS_wNA
            data_TS_wNA_Clusters_h$cluster_labels <- cluster_labels
            file_path <- "data_TS_wNA_Hierarchical_cluster.csv"
            if (file.exists(file_path)) {
               message("File exists. It will be replaced.")
            } else {
               message("File does not exist. A new file will be created.")
            }

           write.csv(data_TS_wNA_Clusters_h, file = file_path, row.names = FALSE)

         }
       },
       {
         # Perform Distribution
         if (option == 3) {
            k <- optimal_clusters
            gmm_result_wNA <- perform_gmm_clustering(data_TS_wNA_scaled, k)
            plot_TS_wNA<-create_pca_plot_gmm(data_TS_wNA_scaled, gmm_result_wNA)
            print(plot_TS_wNA)
            generate_cluster_summary_statistics_gmm(data_TS_wNA, gmm_result_wNA)
            cluster_labels <- gmm_result_wNA$classification
            data_TS_wNA_Clusters_d<-cbind(data_TS_wNA, cluster_labels)
            file_path <- "data_TS_wNA_Distribution_cluster.csv"
            if (file.exists(file_path)) {
                message("File exists. It will be replaced.")
            } else {
                message("File does not exist. A new file will be created.")
            }
           write.csv(data_TS_wNA_Clusters_d, file = file_path, row.names = FALSE)

         }
       }
)




if ("cluster" %in% colnames(dataset)) {
  # If it exists, delete the 'cluster_labels' column
  dataset <- dataset[, -which(colnames(dataset) == "cluster")]
}

# Rename columns, first stays original, rest are numbered sequentially
col_num_name <- seq(1, ncol(dataset)-1, by=1)
col_num_name <- c(names(dataset[1]), col_num_name)
colnames(dataset) <- col_num_name

# Convert the dataset to a matrix and then to a vector, extracting unique columns
unique_tokens <- unique(as.vector(as.matrix(dataset[,2:ncol(dataset)])))

# Assign high-contrast colors to each unique token for heatmap coloring
heatmap_color <- structure(kelly(n=length(unique_tokens)), names = unique_tokens)

print(unique_tokens)

switch(option,
       {
         # Perform K-means Clustering
         if (option == 1) {

            dataset$cluster <- data_TS_wNA_Clusters_k$cluster_labels
            
            for (i in 1:max(dataset$cluster)){
            
              # Subsetting dataset by cluster
              data_clust <- dataset[dataset$cluster == i, ]  
              
              # Remove last column
              data_clust$cluster <- NULL
              
              # Aggregating by token count at each point in the sequence 
              token_states <- data_clust %>% gather(key, value, -UniqueID) %>% dplyr::select(-UniqueID) %>%                    table()                        
            
              # Turning token_states into a dataframe...
              state_map <- as.data.frame.matrix(token_states) 
            
              # Creating a rowname vector and ordering on this (rownames are characters - so convert) 
              rn <- rownames(state_map)
              state_map <- state_map[order(as.numeric(rn)), ]
              
              # Reorders columns
              state_map <- state_map[,order(colSums(-state_map,na.rm=TRUE))]
            
              allnames <- colnames(state_map)
            
              # Creating x-axis numeric label
              state_map$SeqPosition <- as.numeric(rownames(state_map))
            
              # Turn state-map into a data table
              state_map <- setDT(state_map) 
            
              state_final <- melt(state_map, id.vars = "SeqPosition", measure.vars = allnames)
              
              state_final <- rename(state_final, "Token" = "variable")
              
              TS_title <-paste0("Count of token states by sequence position - Cluster: ", as.character(i)) 
            
              # Print state plot
              print(ggplot(state_final, aes(x=SeqPosition, y=value, fill=Token)) + scale_fill_manual(values                    = heatmap_color) + geom_area() + ylab("Token Count") + xlab("Sequence Position") +                               ggtitle(TS_title) + theme(legend.position="bottom"))
              
              # Turning the dataframe into a matrix to use in complexheatmap 
              heatmap_matrix <- as.matrix(data_clust[, 2:(ncol(dataset)-1)]) 
              
              # Setting row/col names to NULL, not necessary
              colnames(heatmap_matrix) = NULL
              rownames(heatmap_matrix) = NULL
              
              # This eliminates an option message
              ht_opt$message = FALSE 
            
              HM_title <-paste0("Token Heatmap Cluster: ", as.character(i))
              
              # Printing heatmap
              print(Heatmap(heatmap_matrix, name = "mat", col = heatmap_color, column_title = HM_title))
            }

          }
            
         },
         {
          if (option == 2) {
            dataset$cluster <- data_TS_wNA_Clusters_h$cluster_labels

            for (i in 1:max(dataset$cluster)){
            
              # Subsetting dataset by cluster
              data_clust <- dataset[dataset$cluster == i, ]  
              
              # Remove last column
              data_clust$cluster <- NULL
              
              # Aggregating by token count at each point in the sequence 
              token_states <- data_clust %>% gather(key, value, -UniqueID) %>% dplyr::select(-UniqueID) %>%                    table()                        
            
              # Turning token_states into a dataframe...
              state_map <- as.data.frame.matrix(token_states) 
            
              # Creating a rowname vector and ordering on this (rownames are characters - so convert) 
              rn <- rownames(state_map)
              state_map <- state_map[order(as.numeric(rn)), ]
              
              # Reorders columns
              state_map <- state_map[,order(colSums(-state_map,na.rm=TRUE))]
            
              allnames <- colnames(state_map)
            
              # Creating x-axis numeric label
              state_map$SeqPosition <- as.numeric(rownames(state_map))
            
              # Turn state-map into a data table
              state_map <- setDT(state_map) 
            
              state_final <- melt(state_map, id.vars = "SeqPosition", measure.vars = allnames)
              
              state_final <- rename(state_final, "Token" = "variable")
              
              TS_title <-paste0("Count of token states by sequence position - Cluster: ", as.character(i)) 
            
              # Print state plot
              print(ggplot(state_final, aes(x=SeqPosition, y=value, fill=Token)) + scale_fill_manual(values                    = heatmap_color) + geom_area() + ylab("Token Count") + xlab("Sequence Position") +                               ggtitle(TS_title) + theme(legend.position="bottom"))
              
              #Turning the dataframe into a matrix to use in complexheatmap 
              heatmap_matrix <- as.matrix(data_clust[, 2:(ncol(dataset)-1)]) 
            
              # Setting row/col names to NULL, not necessary
              colnames(heatmap_matrix) = NULL
              rownames(heatmap_matrix) = NULL
              
              # This eliminates an option message
              ht_opt$message = FALSE 
            
              HM_title <-paste0("Token Heatmap Cluster: ", as.character(i))
              
              # Printing heatmap
              print(Heatmap(heatmap_matrix, name = "mat", col = heatmap_color, column_title = HM_title))
            }
          }
        },
           {
             if (option == 3) {
                dataset$cluster <- data_TS_wNA_Clusters_d$cluster_labels
                for (i in 1:max(dataset$cluster)){
                  data_clust <- dataset[dataset$cluster == i, ]  
                  data_clust$cluster <- NULL
                  token_states <- data_clust %>% gather(key, value, -UniqueID) %>% dplyr::select(-UniqueID)                        %>% table()    
                  state_map <- as.data.frame.matrix(token_states) 
                  rn <- rownames(state_map)
                  state_map <- state_map[order(as.numeric(rn)), ]
                  state_map <- state_map[,order(colSums(-state_map,na.rm=TRUE))]
                  allnames <- colnames(state_map)
                  state_map$SeqPosition <- as.numeric(rownames(state_map))
                  state_map <- setDT(state_map)
                  state_final <- melt(state_map, id.vars = "SeqPosition", measure.vars = allnames)
                  state_final <- rename(state_final, "Token" = "variable")
                  TS_title <-paste0("Count of token states by sequence position - Cluster: ",                                      as.character(i)) 
                  print(ggplot(state_final, aes(x=SeqPosition, y=value, fill=Token)) +                                             scale_fill_manual(values = heatmap_color) + geom_area() + ylab("Token Count") +                                  xlab("Sequence Position") + ggtitle(TS_title) + theme(legend.position="bottom"))
                  
                  # Turning the dataframe into a matrix to use in complexheatmap 
                  heatmap_matrix <- as.matrix(data_clust[, 2:(ncol(dataset)-1)])
                
                  # Setting row/col names to NULL, not necessary
                  colnames(heatmap_matrix) = NULL
                  rownames(heatmap_matrix) = NULL
                  
                  # This eliminates an option message
                  ht_opt$message = FALSE 
                
                  HM_title <-paste0("Token Heatmap Cluster: ", as.character(i))
                  
                  # Printing heatmap
                  print(Heatmap(heatmap_matrix, name = "mat", col = heatmap_color, column_title = HM_title))
                }
               
              }
           }
)
  


detach("package:mclust")

```









